Submission for Peer Assesment 1: Reproducible Research
======================================================

## Loading and preprocessing data

Loading the dataset:

```{r cache=FALSE}
rm(list = as.character(ls()))
setwd("C:\\Misc\\Git\\RepData_PeerAssessment1")

activity_raw <- read.csv(unz("activity.zip",filename="activity.csv"))
```

Cleaning the dataset (removing records with NAs in steps).

```{r cache=FALSE}
activity_cleaned<-activity_raw[!is.na(activity_raw$steps),]
```

## What is mean total number of steps taken per day?

Aggregating to find daily activity:
```{r cache=FALSE}
daily_activity<-aggregate(steps ~ date, data=activity_cleaned, sum)
```

Plotting histogram:
```{r cache=FALSE}
hist(daily_activity$steps, col="Green", breaks=20)
rug(daily_activity$steps)
```

Mean and median of the number of steps take per day:
```{r cache=FALSE}
mean1 <- mean(daily_activity$steps)
```
```{r echo=FALSE, cache=FALSE}
mean1
```
```{r}
median1 <- median(daily_activity$steps)
```
```{r echo=FALSE, cache=FALSE}
median1
```

## What is the average daily activity pattern?

### Calculating average daily activity, and re-coding intervals for better plotting:
```{r cache=FALSE}
interval_activity<-aggregate(steps~interval, data = activity_cleaned, mean)

# re-coding interval consistently for better plotting
interval_activity$i <- seq(0:(dim(interval_activity)[1] - 1))
interval_activity$interval_recoded <- interval_activity$i*5
```

Plotting averaged 5-minute interval:
```{r cache=FALSE}
plot(x=interval_activity$interval_recoded, y=interval_activity$steps,type="l",
     xlab="5-minute interval ", ylab="average number of steps taken")
```

Finding the 5-minute interval with maximum number of steps:
```{r cache=FALSE}
bestInterval <- interval_activity[interval_activity$steps==max(interval_activity$steps),]$interval_recoded

# Calculating time at the beginning of this interval
from_hr<-as.integer(bestInterval/60)
from_min<-bestInterval - as.integer(bestInterval/60)*60
from<-paste(from_hr,from_min, sep=":")

# Calculating time at the end  of this interval
to_hr<-as.integer((bestInterval+5)/60)
to_min<-(bestInterval+5) - as.integer((bestInterval+4)/60)*60
to<-paste(to_hr,to_min, sep=":")
```

Maximum number of steps were taken from `r from` hours (inclusive) to `r to` hours (excluded).  
  

## Imputing missing values

### Calculating total number of missing values

Calculating total number of missing values:
```{r cache=FALSE}
cnt_na <- table(is.na(activity_raw$steps))[["TRUE"]]
```
`r cnt_na` of `r dim(activity_raw)[1]` values in the dataset are missing.

Creating new data set by replacing NAs with mean value for respective interval:

```{r cache=FALSE}
activity_imputed<-activity_raw

for (i in seq(1:dim(activity_imputed)[1]))
{
  if(is.na(activity_imputed[i,]$steps))
  {
    activity_imputed[i,]$steps <- interval_activity[interval_activity$interval == activity_imputed[i,]$interval,]$steps
  }
}
```

Aggregating daily activity from the imputed dataset:

```{r cache=FALSE}
daily_activity_imputed<-aggregate(steps ~ date, data=activity_imputed, sum)
```

Plotting histogram for number of steps taken (from imputed dataset):
```{r cache=FALSE}
hist(daily_activity_imputed$steps, col="Green", breaks=20)
rug(daily_activity_imputed$steps)
```

Mean and median of the number of steps take per day:
```{r cache=FALSE}
mean2 <- mean(daily_activity_imputed$steps)
```
```{r echo=FALSE,  cache=FALSE}
mean2
```
```{r cache=FALSE}
median2 <- median(daily_activity_imputed$steps)
```
```{r echo=FALSE, cache=FALSE}
median2
```

Mean increased by `r mean2-mean1`

Adding factor variable distinguishing weekdays from weekends:
```{r cache=FALSE}
day_indicator<-weekdays(as.Date(activity_imputed$date, "%Y-%m-%d"))
day_indicator[day_indicator=="Sunday" | day_indicator=="Saturday"]<-"weekend"
day_indicator[day_indicator!="weekend"]<-"weekday"
activity_imputed$weekday_type <- as.factor(day_indicator)
```

Carving-out seperate datasets for weekday and weekend activity:
```{r cache=FALSE}
data_wd<-activity_imputed[activity_imputed$weekday_type=="weekday",]
interval_activity_weekday<-aggregate(steps~interval, data = data_wd, mean)
data_we<-activity_imputed[activity_imputed$weekday_type=="weekend",]
interval_activity_weekend<-aggregate(steps~interval, data = data_we, mean)
```


Re-coding interval in weekday and weekend datasets for better plotting:
```{r cache=FALSE}
interval_activity_weekday$i <- seq(1:dim(interval_activity_weekday)[1])
interval_activity_weekday$interval_recoded <- interval_activity_weekday$i*5

interval_activity_weekend$i <- seq(1:dim(interval_activity_weekend)[1])
interval_activity_weekend$interval_recoded <- interval_activity_weekend$i*5
```

Plotting
```{r cache=FALSE}
par(mfrow=c(2,1))

plot(x=interval_activity_weekend$interval_recoded, y=interval_activity_weekend$steps,type="l", xlab="weekend intervals", ylab="Number of Steps")
plot(x=interval_activity_weekday$interval_recoded, y=interval_activity_weekday$steps,type="l", xlab="weekday intervals", ylab="Number of Steps")
```
